Speeding up eQTL scans in the BXD population using GPU

### 2019-08-10 paper.tex
corrected typos and added data info

\noindent The BXD mouse strains are an important set of recombinant
inbred lines with deep phenotypic characterization.  Since the strains
have been genotyped and sequenced, once a set of strains is
phenotyped, the data can be immediately used for QTL (quantitative
trait locus) mapping, and associational analysis with
previously-collected phenotypes.  For ``omic'' phenotypes collected
using high-throughput technologies additional analyses (such as
transcriptional network construction or causal mediation analyses) are
possible.  The GeneNetwork project ({\tt http://gn2.genenetwork.org})
seeks to facilitate this process by providing a searchable database of
phenotypes and genotypes in a variety of organisms (including mouse,
rat, and Arabidopsis).  It provides a suite of interactive tools for
browsing the data, QTL mapping, correlational analyses, network
construction, and genome browsing.  Our goal is to develop a real-time
backend for GeneNetwork to perform eQTL analysis (where we perform
single-trait QTL mapping for thousands of ``omic'' traits) for
important genetic reference populations such as the BXD population.

@@

To perform eQTL scans in the BXD population, one has to simply perform
as many genome scans as there are phenotypes.  This can be done in an
``embarassingly parallel'' fashion by using standard algorithms for
QTL analysis, such as those employed by R/qtl (cite).  In practice,
this is too slow, and substantial speedups are possible.  [Mention
  ideas that have been used before] For example, by using the
Haley-Knott algorithms using genotype probabilities on batches of
phenotypes with the same missing data pattern, instead of using the EM
algorithm on each phenotype individually, substantial speedups are
possible.  This is a well-known trick and is used by R/qtl.
Additionally, if only additive effects are tested, or if the
population has only two genotype categories (as in a backcross or
recombinant inbred line), then matrix multiplication can be used to
perform Haley-Knott regression \citep{shabalin2012matrix}.

@@

\subsection{BXD genotypes and phenotypes}

Describe genotype and phenotype data here.

@@

\subsection{Acceleration Techniques}
We used various techniques to accelerate eQTL scans. 
\subsubsection{Multithreaded CPU operations}
% Used for multi-threaded loops whenever possible, eg. for element-wise operations. 
 Our goal is to build a backend for GeneNetwork that allows researcher to interact with data in real time. 
 Such requirement sets up the standard that the genome scan must finish within seconds.
 To bring out the best performance of CPU, we use multithreaded operations whenever possible. 
 Julia \citep{bezanson2017julia}, our choice of programming language, provide simple yet safe syntax for multithreading. 
 It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multhreaded region. 
 Using \code{Threads.nthreads()} function can show the number of threads in julia, and the default number of threads we use is 16 threads.

@@

% OpenBLAS for multi-threaded matrix operations. 
Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result, regardless of computing platform. 
There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
They target different hardware or use various techniques to get highly optimal results. 
The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
We chose OpenBLAS as CPU computing library.

Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power.
Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
Our chosen hardware for GPU is from Nvidia, which requires their propriatory library CUDA. 
Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
\textit{cuBLAS} \citep{cublas} library is a fast GPU implementation of the BLAS from Nvidia. 
For matrix operations on GPU, we used cuBLAS library.

Not only we want to see how much speed up we can get from using GPU, but also experiment whether the shape of matrix will affect the speedup. 
Of course, most of the time, one can not pick and choose the size and shape of data in matrix form, 
but such information would help researchers as a rough guidance of whether it is worth it to consider the GPU option before investing programming effort for GPU. 
We run matrix multiplication with different shapes matrices, and compared the running time of CPU and GPU. 
CPU time is matrix multiplication from OpenBLAS library using 16 threads. 
GPU time includes all overhead of using GPU, which involves device launch, data transfer and all necessary API calls.
We think to make a fair comparison between CPU and GPU, we need to use maximum strength of both, and include all necessary cost.

The experiment setup is, to multiply two input matrices, A(m$\times$n), and B(n$\times$p), and produce an output matrix C(m$\times$p)
The range of m, n and p is between 2\textsuperscript{4} and 2\textsuperscript{17} in log scale.
Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix total is less than 16 GB. 
The result is shown in Figure \ref{GPUCPUShape}. 
The X axis of Figure \ref{GPUCPUShape} is the dimensions of matrix whose I/O size is over 12 GB.
This is because using GPU involves a lot of overhead. 
Such overhead can only be justified by processing large data sets. 
The Y axis is the speedup of GPU comparing to CPU in log scale.  
From this figure, the matrices whose shape are closer to a square matrix, get better speedup from GPU.
Matrix multiplication is up to 5 times faster on GPU compared to 16 threaded CPU.

@@

\subsubsection{Single precision}
% Faster and memory efficient; may be less accurate, but acceptable for quick scan
Precision means the smallest difference between two representable numbers.
Floating point numbers, in scientific computing, are usually stored in double presicision. 
Double precision floating point numbers takes up 8 bytes in memory, while single precision takes up 4 bytes. 
In addition to the difference in storing size, the speed for calculation using single and double precision also varies depending on hardware. 
For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS) for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80. 
Differences in optimization techniques, underlying architecture cause such performance gap. 
Considering the storage size, data transfer speed and throughput, single precision brings multiple benefits when precision is not the priority concern.

@@

%Specialized computations such as matrix multiplications and element-wise operations; requires additional programming effort
Originally used in graphics, GPU has taken off as a general computing device in recent years because of its massive number of cores at a lower price range, and fast GPGPU libraries such as CUDA and OpenCL.
Based on our profiling result, the time consuming parts of our genome scan method are matrix multiplication and element-wise operations. 
Both are amendable to GPU's heterogeneous computing architecture, since they have no data race conditions and low data dependencies. 
However, GPU also has its own limitations. 
To truely utilize the maximum computing power of GPU, one needs to think creatively to work around such limitations from GPU. 
For example, during our experiemnt, memory transfer between host and device is really slow. 
Profiling result shows that 98\% of total genome scan time is spent on memory transfer. 
To cope with this limitation, instead of off-loading the entire correlation matrix, we use GPU to calculate the max lod score of each phenotype, and output the max.

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.

@@

We used 2 BXD data sets of mouse, spleen and hippocampus data sets to test our method. 
Data from 79 BXD strains for 35556 transcripts from spleens were collected using the Affy Mouse Gene 1.0 ST array. GN accession: GN283. Marker data was available on 7321 markers;
Hippocampus data for 93 strains (70 BXDs) summarized at the exon level for 1,236,087 probes/probe sets.
GeneNetwork Accession Number: GN206
Animals were generated by Drs. Lu, Williams, and colleagues. There were usually one male and one female per strain; strain averages were used as the trait of analysis. All animals were young adults (oldest 160 days).  genotype probabilities were calculated using R/qtl.

@@

\subsection{Benefit of using linear model}
Our method uses linear model, and simplified the genome scan process to basic matrix operations. 
The timing of our method is shown in Table \ref{tab:speedup-table}. 
The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
Getting the same result using the same dataset from R/qtl took about 36.5 seconds.

@@

To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
Compared to CPU only, which took 0.55 seconds, this GPU modification provides 7 times speedup. 
As we mentioned earlier, since we are porting the computation to GPU which took 90\% of total execution time, according to Amdahl's Law, the theoretical maximum speed up is 10 times. 
We got 7 times speed up for genome scan using GPU. 
For hippocampus datasets, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
As a backend service to a website, this wait time is much more preferable than the CPU timing for real time interactions.

\subsection{Software Availability}
The source code for this package is publically available on Bitbucket: \url{https://bitbucket.org/Xiaoqi-hu/lm-gpu}.
This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 
Using Julia language and running \code{test.jl} in \textit{test} folder will run our method using either spleen data and hippocampus data in \textit{data} folder.

This package is developed in Julia language, uses openBLAS and CUDA10.1. 
To use this package, one needs to have an Nvidia GPU in the test machine,
and have Julia language binary, openBLAS and CUDA library installed.

@@

LM-GPU provides a fast backend for GeneNetwork that is powered by GPU. 
It enables researchers to do real-time interaction with the data provided in GeneNetwork. 
Such possibility can help researchers experiment with different parameters, and results in insightful findings. 
LM-GPU takes advantage of GPU's massive parallel computing power and Julia's elegant syntax for quick prototyping and implementing. 
We are able to run genome scan for hippocampus data with over one million phenotypes in three seconds.

>>>>

\noindent The BXD mouse strains are an important set of recombinant
inbred lines with deep phenotypic characterization.  Since the strains
have been genotyped and sequenced, once a set of strains is
phenotyped, the data can be immediately used for QTL (quantitative
trait locus) mapping, and associational analysis with
previously-collected phenotypes.  For ``omic'' phenotypes collected
using high-throughput technologies, additional analyses (such as
transcriptional network construction or causal mediation analyses) are
possible.  The GeneNetwork project ({\tt http://gn2.genenetwork.org})
seeks to facilitate this process by providing a searchable database of
phenotypes and genotypes in a variety of organisms (including mouse,
rat, and Arabidopsis).  It provides a suite of interactive tools for
browsing the data, QTL mapping, correlational analyses, network
construction, and genome browsing.  Our goal is to develop a real-time
backend for GeneNetwork to perform eQTL analysis (where we perform
single-trait QTL mapping for thousands of ``omic'' traits) for
important genetic reference populations such as the BXD population.

@@

To perform eQTL scans in the BXD population, one has to simply perform
as many genome scans as there are phenotypes.  This can be done in an
``embarassingly parallel'' fashion by using standard algorithms for
QTL analysis such as those employed by R/qtl (cite).  In practice,
this is too slow, and substantial speedups are possible.  [Mention
  ideas that have been used before] For example, by using the
Haley-Knott algorithms using genotype probabilities on batches of
phenotypes with the same missing data pattern, instead of using the EM
algorithm on each phenotype individually, substantial speedups are
possible.  This is a well-known trick and is used by R/qtl.
Additionally, if only additive effects are tested, or if the
population has only two genotype categories (as in a backcross or
recombinant inbred line), then matrix multiplication can be used to
perform Haley-Knott regression \citep{shabalin2012matrix}.

@@

% This data subsection is repeated in the last subsection (information and availability of Data sets)
%\subsection{BXD genotypes and phenotypes}

%Describe genotype and phenotype data here.

@@

\subsection{Acceleration Techniques}
We used following techniques to accelerate eQTL scans. 
\subsubsection{Multithreaded CPU operations}
% Used for multi-threaded loops whenever possible, eg. for element-wise operations. 
 Our goal is to build a backend for GeneNetwork that allows researcher to interact with data in real time. 
 Such requirement sets up the standard that the genome scan must finish within seconds.
 To bring out the best performance of CPU, we use multithreaded operations whenever possible. 
 Julia \citep{bezanson2017julia}, our choice of programming language, provide simple yet safe syntax for multithreading. 
 It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multhreaded region. 
 Using \code{Threads.nthreads()} function can show the number of threads in julia, and the default number of threads we use is 16 threads.

@@

% OpenBLAS for multi-threaded matrix operations. 
Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result, regardless of computing platform. 
There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
They target different hardware or use various techniques to get highly optimal results. 
The matrix multiplication in OpenBLAS is default multithreaded and therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
We chose OpenBLAS as CPU computing library.

Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, so that they are amendable to GPU's parallel computing power.
Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
Our chosen hardware for GPU is from Nvidia, which requires its propriatory library CUDA. 
Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
\textit{cuBLAS} \citep{cublas} library is a fast GPU implementation of the BLAS from Nvidia. 
For matrix operations on GPU, we used cuBLAS library.

Not only do we want to see how much speed up we can get from using GPU, but we also experiment whether the shape of matrix will affect the speedup. 
Of course, most of the time, one can not pick the size and shape of data in a matrix form, 
but such information would help researchers as a rough guidance of whether it is worth considering the GPU option before investing programming effort for GPU. 
We run matrix multiplication with different shapes of matrices, and compared the running time of CPU and GPU. 
CPU time is measured in matrix multiplication from OpenBLAS library using 16 threads. 
GPU time includes all overhead of using GPU, which involves device launch, data transfer and all necessary API calls.
In order to make a fair comparison between CPU and GPU, we need to use maximum strength of both and include all necessary cost.

The experiment setup is to multiply two input matrices, A(m$\times$n), and B(n$\times$p), and produce an output matrix C(m$\times$p).  
The range of m, n and p is between 2\textsuperscript{4} and 2\textsuperscript{17} in log scale.
Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix, in total, is less than 16 GB. 
The result is shown in Figure \ref{GPUCPUShape}. 
The X axis of Figure \ref{GPUCPUShape} is the dimensions of matrix whose I/O size is over 12 GB.
This is because using GPU involves a lot of overhead. 
Such overhead can only be justified by processing large data sets. 
The Y axis is the speedup of GPU compared to CPU in log scale.  
From this figure, the matrices whose shapes are closer to square matrices get better speedups from GPU.
Matrix multiplication is up to 5 times faster on GPU compared to that on 16 threaded CPU.

@@

\subsubsection{Single precision}
% Faster and memory efficient; may be less accurate, but acceptable for quick scan
Precision means the smallest difference between two representable numbers.
Floating point numbers, in scientific computing, are usually stored in double presicision. 
Double precision takes up 8 bytes in memory, while single precision takes up 4 bytes. 
In addition to the difference in storing size, the speed for calculation using single and double precisions also varies 
depending on the hardware.  For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS)
for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80. 
Differences in optimization techniques, underlying architecture cause such performance gap. 
Considering the storage size, data transfer speed and throughput, single precision brings multiple benefits when precision is not the priority concern.

@@

%Specialized computations such as matrix multiplications and element-wise operations; requires additional programming effort
Originally used in graphics, GPU has taken off as a general computing device in recent years because of its massive number of cores at a lower price range, and fast GPGPU libraries such as CUDA and OpenCL.
Based on our profiling result, the time consuming parts of our genome scan method are matrix multiplication and element-wise operations. 
Both are amendable to GPU's heterogeneous computing architecture, since they have no data race conditions and low data dependencies. 
However, GPU also has its own limitations. 
To truely utilize the maximum computing power of GPU, one needs to think creatively to work around such limitations from GPU. 
For example, during our experiemnt, memory transfer between host and device is really slow. 
Profiling result shows that 98\% of total genome scan time is spent on memory transfer. 
To cope with this limitation, instead of off-loading the entire correlation matrix, we use GPU to calculate the max lod score of each phenotype and output the max.

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.

@@

We downloaded BXD Genotype Database (GN Accession: GN600) and two sets of gene expression data, 
UTHSC Affy MoGene 1.0 ST Spleen (GN Accession : GN283)
and  UMUTAffy Hippocampus Exon (GN Accession: GN206) from GeneNetwork to test our method.  
The genotype file includes 7321 markers by 198 BXD strains, the spleen consists of 79 BXD strains by 35556 transcripts, 
and the hippocampus does 70 BXD strains by 1,236,087 probes/probe sets.  
Genotypes were subsetted by strains, pseudo markers were added only for the hippocampus data, 
and genotype probabilities were computed accordingly.  We thus used genotypes of 79 strains by 7321 markers for the spleen data, and 
those of 70 by 8570 for the hippocampus data for eQTL scan.  The whole cleaning process was performed using R/qtl, R/qtl2.

@@

\subsection{Benefit of using linear model}
Our method uses the linear model and simplified the genome scan process to basic matrix operations. 
The timing of our method is shown in Table \ref{tab:speedup-table}. 
The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
Getting the same result using the same dataset from R/qtl took about 36.5 seconds.

@@

To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method because main interest of genome scan is to find the maximum LOD score of every phenotype. 
This step is highly parallelizable, can utilize GPU's massive cores, and, at the same time, reduce the amount of data that needs to be transfered back to host. 
Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
Compared to CPU only, which took 0.55 seconds, this GPU modification provides 7 times speedup. 
As we mentioned earlier, since we are porting the computation to GPU which took 90\% of total execution time, according to Amdahl's Law, the theoretical maximum speed up is 10 times. 
We got 7 times speed up for genome scan using GPU. 
For the hippocampus dataset, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
As a backend service to a website, this wait time is much more preferable than the CPU timing for real time interactions.

\subsection{Software Availability}
The source code for this package is publically available on Bitbucket: \url{https://bitbucket.org/Xiaoqi-hu/lm-gpu}.
This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 
Using Julia language and running \code{test.jl} in \textit{test} folder will run our method using either spleen data or hippocampus data in \textit{data} folder.

This package is developed in Julia language, uses openBLAS and CUDA10.1. 
To use this package, one needs to have a Nvidia GPU in the test machine,
and have Julia language binary, openBLAS and CUDA library installed.

@@

LM-GPU provides a fast backend for GeneNetwork that is powered by GPU. 
It enables researchers to do real-time interaction with the data provided in GeneNetwork. 
Such possibility can help researchers experiment with different parameters and results in insightful findings. 
LM-GPU takes advantage of GPU's massive parallel computing power and Julia's elegant syntax for quick prototyping and implementing. 
We are able to run genome scan for hippocampus data with over one million phenotypes in three seconds.
